Limited plasticity but increased variance in physiological rates across ectotherm populations under climate change

### 2025-01-06 ms.qmd
Some important additions from Shinichi

1.	Climate change causes warmer and more variable temperatures globally, impacting physiological rates and function in ectothermic animals. Acclimation of physiological rates can help maintain function. However, it is unresolved how variance in physiological rates changes with temperature despite its potential ecological and evolutionary importance. 
2.	We developed new effect sizes that capture how both the mean and variation in physiological rates change across temperature (based on the temperature coefficient, $Q_{10}$), and used them to test how acclimation and acute thermal responses vary across aquatic and terrestrial ectotherms using meta-analysis (>1900 effects from 226 species). Comparing both the magnitude of acclimation and changes in variation side-by-side provides unique opportunities for evaluating the importance of plasticity and selection under climate change.
3.	We show that variance in physiological rates increases at higher temperatures, but that the magnitude of change depends on habitat. Freshwater and marine ectotherms are capable of acclimation and have the greatest increase in variance. In contrast, terrestrial ectotherms lack the capacity for acclimation and have smaller increases in variance. Simulations suggest that these patterns may result from differences in among-individual variation in thermal breadth and optima of performance curves across habitats. 
4.	Our results highlight the greater vulnerability of terrestrial ectotherms to climate change because of both a lack of acclimation capacity and a limited increases in variance that may provide less raw material for evolutionary adaptation. Considering both acclimation capacity and variance in physiological rates side-by-side is therefore important for understanding how climate change will impact populations.

@@

where, $s_{lnRR_{Q_{10_{acute}}}}$ and $s_{lnRR_{Q_{10_{acclimation}}}}$ is the sampling variance for the acute and acclimation effect sizes, respectively. Again, all notation is defined in @eq-Vlnq10 in the main manuscript.

@@

```{r, fig-delta}
#| label: fig-delta
#| fig-cap: "Meta-analysis results for $\\Delta lnRR_{Q_{10}}$ across different habitats. Thick bars are 95% confidence intervals (CI) and thin bars 95% prediction intervals (PI). $\\mu$ values are the overall meta-analytic means averaged across acute and acclimation types within each habitat type. 95% CI's are indicated within square brackets and raw effects are weighted by their precision (inverse sampling variance). $p_{MCMC}$ values are the posterior probability of the contrast or overall meta-analytic mean being different from zero. k = total number of effect size estimates while the numbers in brackets indicate the number of species.  For ease of visualisation, all the raw data plotted for both acute and acclimation type effect sizes are presented as circles. Raw effects are weighted by their precision (inverse sampling variance). "

>>>>

1.	Climate change causes warmer and more variable temperatures globally, impacting physiological rates and function in ectothermic animals. Acclimation of physiological rates can help maintain function. However, it is unresolved how variance in physiological rates changes with temperature despite its potential ecological and evolutionary importance. 
2.	We developed new effect sizes that capture how both the mean and variation in physiological rates change across temperature (based on the temperature coefficient, $Q_{10}$), and used them to test how acclimation and acute thermal responses vary across aquatic and terrestrial ectotherms using meta-analysis (>1900 effects from 226 species). Comparing both the magnitude of acclimation and changes in variation side-by-side provides unique opportunities for evaluating the importance of plasticity and selection under climate change.
3.	We show that variance in physiological rates increases at higher temperatures, but that the magnitude of change depends on habitat. Freshwater and marine ectotherms are capable of acclimation and have the greatest increase in variance. In contrast, terrestrial ectotherms lack the capacity for acclimation and have smaller increases in variance. Simulations suggest that these patterns may result from differences in among-individual variation in thermal breadth and optima of performance curves across habitats. 
4.	Our results highlight the greater vulnerability of terrestrial ectotherms to climate change because of both a lack of acclimation capacity and a limited increase in variance that may provide less raw material for evolutionary adaptation. Considering both acclimation capacity and variance in physiological rates side-by-side is therefore important for understanding how climate change will impact populations.

@@

where, $s_{lnRR_{Q_{10_{acute}}}}$ and $s_{lnRR_{Q_{10_{acclimation}}}}$ is the sampling variance for the acute and acclimation effect sizes, respectively (note that we assume the independence of ‘acute’ and ‘acclimation’ groups). Again, all notation is defined in @eq-Vlnq10 in the main manuscript.

@@

```{r, fig-delta}
#| label: fig-delta
#| fig-cap: "Meta-analysis results for $\\Delta lnRR_{Q_{10}}$ across different habitats. Thick bars are 95% confidence intervals (CI) and thin bars 95% prediction intervals (PI). $\\mu$ values are the overall meta-analytic means averaged across acute and acclimation types within each habitat type. 95% CI's are indicated within square brackets and raw effects are weighted by their precision (inverse sampling variance). $p_{MCMC}$ values are the posterior probability of the contrast or overall meta-analytic mean being different from zero. k = total number of effect size estimates, while the numbers in brackets indicate the number of species.  For ease of visualisation, all the raw data plotted for both acute and acclimation-type effect sizes are presented as circles. Raw effects are weighted by their precision (inverse sampling variance). "
